FOODIE, Fortran Object oriented Ordinary Differential Equations integration library based on Abstract Calculus Pattern

### 2015-10-19 papers/announcement/src/introduction/introduction.tex
Change project name to FOODIE
FOODIE => Fortran Object-Oriented Differential-equations Integration Environment

Why:

Due to a great idea of Damian.

This change addresses the need by:

Change all references to old FOODiE acronym.

Side effects:

A lot...

FOODiE library is designed for solving the generalized IVP \ref{eq:IVP}, it being completely unaware of the actual problem's definition. FOODiE library provides a high-level, well-documented, simple Application Program Interface (API) for many well-known ODE integration solvers, its aims being twofold:

@@

FOODiE library must:

@@

FOODiE, meaning Fortran Object oriented Ordinary Differential Equations integration library, has been developed with the aim to satisfy the above specifications. The present paper is its first comprehensive presentation.

The Fortran (Formula Translator) programming language is the \emph{de facto} standard into computer science field: it strongly facilitates the effective and efficient translation of (even complex) mathematical and numerical models into an operative software without compromise on computations speed and accuracy. Moreover, its simple syntax is suitable for scientific researchers that are interested (and skilled) in the physical aspects of the numerical computations rather than computer technicians. Consequently, we develop FOODiE using Fortran language: FOODiE is written by research scientists for research scientists.

One key-point of the FOODiE development is the \emph{problem generalization}: the problem solved must be the IVP \ref{eq:IVP} rather than any of its actual definitions. Consequently, we must rely on a \emph{generic} implementation of the solvers. To this aim, OOP is very useful: it allows to express IVP \ref{eq:IVP} in a very concise and clear formulation that is really generic. In particular, our implementation is based on \emph{Abstract Calculus Pattern} (ACP) concept.

\paragraph{The Abstract Calculus Pattern}
The abstract calculus pattern provides a simple solution for the connection between the very high-level expression of IVP \ref{eq:IVP} and the eventual concrete (low-level) implementation of the ODE problem being solved. ACP essentially constitutes a \emph{contract} based on an Abstract Data Type (ADT): we specify an ADT supporting a certain set of mathematical operators (differential and integral ones) and implement FOODiE solvers only on the basis of this ADT. FOODiE clients must formulate the ODE problem under integration defining their own concrete extensions of our ADT (implementing all the deferred operators). Such an approach defines the abstract calculus pattern: FOODiE solvers are aware of only the ADT, while FOODiE clients extend the ADT for defining the concrete ODE problem.

Is is worth noting that this ACP emancipates the solvers implementations from any low-level problem-dependent details: the ODE solvers developed with this pattern are extremely concise, clear, maintainable and less errors-prone with respect a low-level (non abstract) pattern. Moreover, the FOODiE clients can use solvers being extremely robust: as a matter of facts, FOODiE solvers are expressed in a formulation very close to the mathematical one and are tested on an extremely varying family of problems. As shown in the following, such a great flexibility does not compromise the computational efficiency.

The present paper is organized as following: in section \ref{sec:MNmodels} a brief description of the mathematical and numerical methods currently implemented into FOODiE is presented; in section \ref{sec:API} a detailed discussion on the implementation specifications is provided by means of an analytical code-listings review; in section \ref{sec:Tests} a verification analysis on the results of FOODiE applications is presented; section \ref{sec:parallel} provides an analysis of FOODiE performances under parallel frameworks scenario like the OpenMP and MPI paradigms; finally, in section \ref{sec:conclusions} concluding remarks and perspectives are depicted.

>>>>

FOODIE library is designed for solving the generalized IVP \ref{eq:IVP}, it being completely unaware of the actual problem's definition. FOODIE library provides a high-level, well-documented, simple Application Program Interface (API) for many well-known ODE integration solvers, its aims being twofold:

@@

FOODIE library must:

@@

FOODIE, meaning Fortran Object oriented Ordinary Differential Equations integration library, has been developed with the aim to satisfy the above specifications. The present paper is its first comprehensive presentation.

The Fortran (Formula Translator) programming language is the \emph{de facto} standard into computer science field: it strongly facilitates the effective and efficient translation of (even complex) mathematical and numerical models into an operative software without compromise on computations speed and accuracy. Moreover, its simple syntax is suitable for scientific researchers that are interested (and skilled) in the physical aspects of the numerical computations rather than computer technicians. Consequently, we develop FOODIE using Fortran language: FOODIE is written by research scientists for research scientists.

One key-point of the FOODIE development is the \emph{problem generalization}: the problem solved must be the IVP \ref{eq:IVP} rather than any of its actual definitions. Consequently, we must rely on a \emph{generic} implementation of the solvers. To this aim, OOP is very useful: it allows to express IVP \ref{eq:IVP} in a very concise and clear formulation that is really generic. In particular, our implementation is based on \emph{Abstract Calculus Pattern} (ACP) concept.

\paragraph{The Abstract Calculus Pattern}
The abstract calculus pattern provides a simple solution for the connection between the very high-level expression of IVP \ref{eq:IVP} and the eventual concrete (low-level) implementation of the ODE problem being solved. ACP essentially constitutes a \emph{contract} based on an Abstract Data Type (ADT): we specify an ADT supporting a certain set of mathematical operators (differential and integral ones) and implement FOODIE solvers only on the basis of this ADT. FOODIE clients must formulate the ODE problem under integration defining their own concrete extensions of our ADT (implementing all the deferred operators). Such an approach defines the abstract calculus pattern: FOODIE solvers are aware of only the ADT, while FOODIE clients extend the ADT for defining the concrete ODE problem.

Is is worth noting that this ACP emancipates the solvers implementations from any low-level problem-dependent details: the ODE solvers developed with this pattern are extremely concise, clear, maintainable and less errors-prone with respect a low-level (non abstract) pattern. Moreover, the FOODIE clients can use solvers being extremely robust: as a matter of facts, FOODIE solvers are expressed in a formulation very close to the mathematical one and are tested on an extremely varying family of problems. As shown in the following, such a great flexibility does not compromise the computational efficiency.

The present paper is organized as following: in section \ref{sec:MNmodels} a brief description of the mathematical and numerical methods currently implemented into FOODIE is presented; in section \ref{sec:API} a detailed discussion on the implementation specifications is provided by means of an analytical code-listings review; in section \ref{sec:Tests} a verification analysis on the results of FOODIE applications is presented; section \ref{sec:parallel} provides an analysis of FOODIE performances under parallel frameworks scenario like the OpenMP and MPI paradigms; finally, in section \ref{sec:conclusions} concluding remarks and perspectives are depicted.

### 2015-10-19 papers/announcement/src/mathematical-numerical-models/mathematical-numerical-models.tex
Change project name to FOODIE
FOODIE => Fortran Object-Oriented Differential-equations Integration Environment

Why:

Due to a great idea of Damian.

This change addresses the need by:

Change all references to old FOODiE acronym.

Side effects:

A lot...

In many (most) circumstances, the solution of equation \ref{eq:IVP} cannot be computed in a closed, exact form (even if it exists and is unique) due to the complexity and nature of the residuals functions, that is often non linear. Consequently, the problem is often solved relying on a numerical approach: the solution of system \ref{eq:IVP} at a time $t^n$, namely $U(t^n)$, is approximated by a subsequent time-marching approximations $U_0=u_0 \rightarrow u_1 \rightarrow u_2 \rightarrow ... \rightarrow u_N\approx U(t^n)$ where the relation $u_i \rightarrow u_{i+1}$ implies a \emph{stepping, numerical integration} from the time $t^i$ to time $t^{i+1}$ and $N$ is the total number of numerical time steps necessary to evolve the initial conditions toward the searched solution $U(t^n)$. To this aim, many numerical schemes have been devised. Notably, the numerical schemes of practical usefulness must posses some necessary proprieties such as \emph{consistency} and \emph{stability} to ensure that the numerical approximation \emph{converges} to the exact solution as the numerical time step tends to zero. A detailed discussion of these details is out the scope of the present work and is omitted. Here, we briefly recall some classifications necessary to introduce the schemes implemented into the FOODiE library.

@@

Essentially, the multi-step schemes have been developed to obtain an accurate approximation of the subsequent numerical steps using the informations contained into the previously computed steps, thus this approach relates the next step approximation to a set of the previously computed steps. On the contrary, a one-step scheme evolves the solution toward the next step using only the information coming from the current time approximation. In the framework of one-step schemes family an equivalent accurate approximation can be obtained by means of a multi-stage approach as the one due to Runge-Kutta. The current version of FOODiE provides schemes belonging to both these families.

The other ODE solvers classification concerns with explicit or implicit nature of the schemes employed. Briefly, an explicit scheme computes the next step approximation using the previously computed steps at most to the current time, whereas an implicit scheme uses also the next step approximation (that is the unknown), thus it requires extra computations. The implicit approach is of practical use for \emph{stiff} systems where the usage of explicit schemes could require an extremely small time step to evolve in a \emph{stable} way the solution. Mixing together explicit and implicit schemes it is possible to build a family of \emph{predictor-corrector} methods: using an explicit scheme to predict a guess for the next step approximation it is possible to use an implicit method for correcting this guess. Currently, FOODiE provides explicit solvers and predictor-correct ones.

FOODiE currently implements the following ODE schemes:

@@

In general, the TVD/SSP Runge-Kutta schemes provided by FOODiE library are written by means of the following algorithm:
\begin{equation}
  U^{n+1} = U^{n} + \Delta t \cdot\sum\limits_{s=1}^{N_{s}}{\beta_s K_s}
\label{eq:RK}
\end{equation}
where $N_s$ is the number of Runge-Kutta stages used and $K_s$ is the $s^{th}$ stage defined as:
\begin{equation}
  K_s = R\left(t^n + \gamma_s \Delta t, U^n+\Delta t\sum\limits_{l=1}^{s-1}{\alpha_{s,l} K_l} \right)
\label{eq:RK-stage}
\end{equation}
It is worth noting that the equations \ref{eq:RK} and \ref{eq:RK-stage} contain also implicit schemes. A scheme belonging to this family is operative once the coefficients $\alpha,\, \beta,\, \gamma$ are provided. We represent these coefficients using the Butcher's table, that for an explicit scheme where $\gamma_1=\alpha_{1,*}=\alpha_{i,i}=0$ has the form reported in table \ref{tab:butcher-table}.

@@

The nature of the scheme and the properties of the solutions obtained depend on the number of stages and on the value of the coefficients selected. Currently, FOODiE provides 3 Runge-Kutta schemes having TVD or Strong Stability Preserving (SSP) propriety (thus they being suitable for ODE systems involving rapidly changing non linear dynamics) the Butcher's coefficients of which are reported in tables \ref{tab:RK-TVD-butcher-2}, \ref{tab:RK-SSP-butcher-3}, \ref{tab:RK-SSP-butcher-5}.

@@

As aforementioned, standard Runge-Kutta schemes have the drawback to require $N_S$ auxiliary memory registers to store the necessary stages data. In order to make an efficient use of the available limited computer memory, the class of low storage Runge-Kutta scheme was devised. Essentially, the standard Runge-Kutta class (under some specific conditions) can be reformulated allowing a more efficient memory management. Currently FOODiE provides a class of \emph{2N registers} storage Runge-Kutta schemes, meaning that the storage of all stages requires only 2 registers of memory with a \emph{word} length $N$ (namely the length of the state vector) in contrast to the standard formulation where $N_s$ registers of the same length $N$ are required. This is a dramatic improvement of memory efficiency especially for schemes using a high number of stages ($N_s \ge 4$) where the memory necessary is an half with respect the original formulation. Unfortunately, not all standard Runge-Kutta schemes can be reformulated as a low storage one.

@@

Currently FOODiE provides 5/6/7/12/13/14 stages, all $4^{th}$ order, 2N registers explicit schemes, the coefficients of which are listed in table \ref{tab:RK-ls}.

@@

In general, the Adams-Bashforth schemes provided by FOODiE library are written by means of the following algorithm (for only explicit schemes):

@@

Currently FOODiE provides 2, 3, and 4 steps schemes having $2^{nd}$, $3^{rd}$ and $4^{th}$ formal order of accuracy, respectively. The $b_s$ coefficients are reported in table \ref{tab:AB-coeff}.

@@

In general, the Adams-Moulton schemes provided by FOODiE library are written by means of the following algorithm (for only implicit schemes):

@@

Currently FOODiE provides 1, 2, and 3 steps schemes having $2^{nd}$, $3^{rd}$ and $4^{th}$ formal order of accuracy, respectively. The $b_s$ coefficients are reported in table \ref{tab:AM-coeff}.

@@

In general, the Adams-Bashforth-Moulton schemes provided by FOODiE library are written by means of the following algorithm:

@@

Currently FOODiE provides $N_s^c=1, 2, 3 \rightarrow N_s^c=2, 3, 4$ steps schemes having $2^{nd}$, $3^{rd}$ and $4^{th}$ formal order of accuracy, respectively. The $b_s^{p,c}$ coefficients are those reported in tables \ref{tab:AB-coeff} and \ref{tab:AM-coeff}.

@@

The unfiltered leapfrog scheme provided by FOODiE is:
\begin{equation}
  U\left(t^{n+2}\right) = U\left(t^{n}\right) + 2\Delta t \cdot R\left[t^{n+1}, U\left(t^{n+1}\right)\right]
\label{eq:leapfrog}
\end{equation}

FOODiE provides, in a \emph{seamless} API, also filtered leapfrog schemes. A widely used filter is due to Robert and Asselin, that suppress the computational modes at the cost of accuracy reduction resulting into a $1^{st}$ order error in amplitude value. A more accurate filter, able to provide a $3^{rd}$ order error on amplitude, is a modification of the Robert-Asselin filter due to Williams known as Robert-Asselin-Williams (RAW) filter, that filters the approximation of $U\left(t^{n+1}\right)$ and $U\left(t^{n+2}\right)$ by the following scalar coefficient:

@@

The filter coefficients should be taken as $\nu \in (0,1]$ and $\alpha \in (0.5,1]$. If $\alpha=0.5$ the filters of time $t^{n+1}$ and $t^{n+2}$ have the same amplitude and opposite sign thus allowing to the optimal $3^{rd}$ order error on amplitude. The default values of the FOODiE provided scheme are $\nu=0.01$ $\alpha=0.53$, but they can be customized at runtime.

>>>>

In many (most) circumstances, the solution of equation \ref{eq:IVP} cannot be computed in a closed, exact form (even if it exists and is unique) due to the complexity and nature of the residuals functions, that is often non linear. Consequently, the problem is often solved relying on a numerical approach: the solution of system \ref{eq:IVP} at a time $t^n$, namely $U(t^n)$, is approximated by a subsequent time-marching approximations $U_0=u_0 \rightarrow u_1 \rightarrow u_2 \rightarrow ... \rightarrow u_N\approx U(t^n)$ where the relation $u_i \rightarrow u_{i+1}$ implies a \emph{stepping, numerical integration} from the time $t^i$ to time $t^{i+1}$ and $N$ is the total number of numerical time steps necessary to evolve the initial conditions toward the searched solution $U(t^n)$. To this aim, many numerical schemes have been devised. Notably, the numerical schemes of practical usefulness must posses some necessary proprieties such as \emph{consistency} and \emph{stability} to ensure that the numerical approximation \emph{converges} to the exact solution as the numerical time step tends to zero. A detailed discussion of these details is out the scope of the present work and is omitted. Here, we briefly recall some classifications necessary to introduce the schemes implemented into the FOODIE library.

@@

Essentially, the multi-step schemes have been developed to obtain an accurate approximation of the subsequent numerical steps using the informations contained into the previously computed steps, thus this approach relates the next step approximation to a set of the previously computed steps. On the contrary, a one-step scheme evolves the solution toward the next step using only the information coming from the current time approximation. In the framework of one-step schemes family an equivalent accurate approximation can be obtained by means of a multi-stage approach as the one due to Runge-Kutta. The current version of FOODIE provides schemes belonging to both these families.

The other ODE solvers classification concerns with explicit or implicit nature of the schemes employed. Briefly, an explicit scheme computes the next step approximation using the previously computed steps at most to the current time, whereas an implicit scheme uses also the next step approximation (that is the unknown), thus it requires extra computations. The implicit approach is of practical use for \emph{stiff} systems where the usage of explicit schemes could require an extremely small time step to evolve in a \emph{stable} way the solution. Mixing together explicit and implicit schemes it is possible to build a family of \emph{predictor-corrector} methods: using an explicit scheme to predict a guess for the next step approximation it is possible to use an implicit method for correcting this guess. Currently, FOODIE provides explicit solvers and predictor-correct ones.

FOODIE currently implements the following ODE schemes:

@@

In general, the TVD/SSP Runge-Kutta schemes provided by FOODIE library are written by means of the following algorithm:
\begin{equation}
  U^{n+1} = U^{n} + \Delta t \cdot\sum\limits_{s=1}^{N_{s}}{\beta_s K_s}
\label{eq:RK}
\end{equation}
where $N_s$ is the number of Runge-Kutta stages used and $K_s$ is the $s^{th}$ stage defined as:
\begin{equation}
  K_s = R\left(t^n + \gamma_s \Delta t, U^n+\Delta t\sum\limits_{l=1}^{s-1}{\alpha_{s,l} K_l} \right)
\label{eq:RK-stage}
\end{equation}
It is worth noting that the equations \ref{eq:RK} and \ref{eq:RK-stage} contain also implicit schemes. A scheme belonging to this family is operative once the coefficients $\alpha,\, \beta,\, \gamma$ are provided. We represent these coefficients using the Butcher's table, that for an explicit scheme where $\gamma_1=\alpha_{1,*}=\alpha_{i,i}=0$ has the form reported in table \ref{tab:butcher-table}.

@@

The nature of the scheme and the properties of the solutions obtained depend on the number of stages and on the value of the coefficients selected. Currently, FOODIE provides 3 Runge-Kutta schemes having TVD or Strong Stability Preserving (SSP) propriety (thus they being suitable for ODE systems involving rapidly changing non linear dynamics) the Butcher's coefficients of which are reported in tables \ref{tab:RK-TVD-butcher-2}, \ref{tab:RK-SSP-butcher-3}, \ref{tab:RK-SSP-butcher-5}.

@@

As aforementioned, standard Runge-Kutta schemes have the drawback to require $N_S$ auxiliary memory registers to store the necessary stages data. In order to make an efficient use of the available limited computer memory, the class of low storage Runge-Kutta scheme was devised. Essentially, the standard Runge-Kutta class (under some specific conditions) can be reformulated allowing a more efficient memory management. Currently FOODIE provides a class of \emph{2N registers} storage Runge-Kutta schemes, meaning that the storage of all stages requires only 2 registers of memory with a \emph{word} length $N$ (namely the length of the state vector) in contrast to the standard formulation where $N_s$ registers of the same length $N$ are required. This is a dramatic improvement of memory efficiency especially for schemes using a high number of stages ($N_s \ge 4$) where the memory necessary is an half with respect the original formulation. Unfortunately, not all standard Runge-Kutta schemes can be reformulated as a low storage one.

@@

Currently FOODIE provides 5/6/7/12/13/14 stages, all $4^{th}$ order, 2N registers explicit schemes, the coefficients of which are listed in table \ref{tab:RK-ls}.

@@

In general, the Adams-Bashforth schemes provided by FOODIE library are written by means of the following algorithm (for only explicit schemes):

@@

Currently FOODIE provides 2, 3, and 4 steps schemes having $2^{nd}$, $3^{rd}$ and $4^{th}$ formal order of accuracy, respectively. The $b_s$ coefficients are reported in table \ref{tab:AB-coeff}.

@@

In general, the Adams-Moulton schemes provided by FOODIE library are written by means of the following algorithm (for only implicit schemes):

@@

Currently FOODIE provides 1, 2, and 3 steps schemes having $2^{nd}$, $3^{rd}$ and $4^{th}$ formal order of accuracy, respectively. The $b_s$ coefficients are reported in table \ref{tab:AM-coeff}.

@@

In general, the Adams-Bashforth-Moulton schemes provided by FOODIE library are written by means of the following algorithm:

@@

Currently FOODIE provides $N_s^c=1, 2, 3 \rightarrow N_s^c=2, 3, 4$ steps schemes having $2^{nd}$, $3^{rd}$ and $4^{th}$ formal order of accuracy, respectively. The $b_s^{p,c}$ coefficients are those reported in tables \ref{tab:AB-coeff} and \ref{tab:AM-coeff}.

@@

The unfiltered leapfrog scheme provided by FOODIE is:
\begin{equation}
  U\left(t^{n+2}\right) = U\left(t^{n}\right) + 2\Delta t \cdot R\left[t^{n+1}, U\left(t^{n+1}\right)\right]
\label{eq:leapfrog}
\end{equation}

FOODIE provides, in a \emph{seamless} API, also filtered leapfrog schemes. A widely used filter is due to Robert and Asselin, that suppress the computational modes at the cost of accuracy reduction resulting into a $1^{st}$ order error in amplitude value. A more accurate filter, able to provide a $3^{rd}$ order error on amplitude, is a modification of the Robert-Asselin filter due to Williams known as Robert-Asselin-Williams (RAW) filter, that filters the approximation of $U\left(t^{n+1}\right)$ and $U\left(t^{n+2}\right)$ by the following scalar coefficient:

@@

The filter coefficients should be taken as $\nu \in (0,1]$ and $\alpha \in (0.5,1]$. If $\alpha=0.5$ the filters of time $t^{n+1}$ and $t^{n+2}$ have the same amplitude and opposite sign thus allowing to the optimal $3^{rd}$ order error on amplitude. The default values of the FOODIE provided scheme are $\nu=0.01$ $\alpha=0.53$, but they can be customized at runtime.

### 2015-10-19 papers/announcement/src/tests/oscillation/oscillation-API.tex
Change project name to FOODIE
FOODIE => Fortran Object-Oriented Differential-equations Integration Environment

Why:

Due to a great idea of Damian.

This change addresses the need by:

Change all references to old FOODiE acronym.

Side effects:

A lot...

\subsubsection{FOODiE aware implementation of an oscillation numerical solver}

The IVP \ref{eq:oscillation} can be easily solved by means of FOODiE library. The first block of a FOODiE aware solution consists to define an \emph{oscillation integrand field} defining a concrete extension of the FOODiE \emph{integrand} type. Listing \ref{list:oscillation_type} reports the implementation of such an integrand field that is contained into the tests suite shipped within the FOODiE library.

@@

As the listing \ref{list:oscillation_type} shows, the FOODiE implementation strictly corresponds to the mathematical formulation embracing all the relevant mathematical aspects into one derived type, a single \emph{object}. Here we not review the implementation of all deferred methods, this being out of the scope of the present work: the interested reader can see the tests suite sources shipped within the FOODiE library. However, some of these methods are relevant for our discussion, thus they are reviewed.

@@

Probably, the most important methods for an IVP solver is the residuals function. As a matter of facts, the ODE equations are implemented into the residuals function. However, the FOODiE ADT strongly alleviates the subtle problems that could arise when the ODE solver is hard-implemented within the specific ODE equations. As a matter of facts, the \emph{integrand} ADT specifies the precise interface the residuals function must have: if the client code implements a compliant interface, the FOODiE solvers will work as expected, reducing the possible errors location into the ODE equations, having designed the solvers on the ADT and not on the concrete type.

@@

The assignment overloading of the \emph{oscillation} type is the last key-method that enforces the conciseness of the FOODiE aware implementation. Listing \ref{list:oscillation_assign} reports the implementation of the assignment overloading. Essentially, to all the data members of the left-hand-side are assigned the values of the corresponding right-hand-side. Notably, for the assignment of the state vector and of the previous time steps solution array we take advantage of the automatic re-allocation of the left-hand-side variables when they are not allocated or allocated differently from the right-hand-side, that is a Fortran 2003 feature. In spite its simplicity, the assignment overloading is a key-method enabling the usage of FOODiE solver: effectively, the assignment between two \emph{integrand} ADT variables is ubiquitous into the solvers implementations, see equation \ref{eq:RK} for example.

@@

\paragraph{FOODiE numerical integration}

Using the above discussed \emph{oscillation} type it is very easy to solve IVP \ref{eq:oscillation} by means of FOODiE library. Listing \ref{list:oscillation_leapfrog} presents the numerical integration of system \ref{eq:oscillation} by means of the Leapfrog RAW-filtered method. In the example, the integration is performed with $10^4$ steps with a fixed $\Delta t=10^2$ until the time $t=10^6$ is reached. The example shows also that for starting a multi-step scheme such as the Leapfrog one a lower-oder or equivalent order one-scheme is necessary: in the example the first 2 steps are computed by means of one-step TVD/SSP Runge-Kutta 2-stages schemes. Note that the memory registers for storing the Runge-Kutta stages and the RAW filter displacement must be handled by the client code. Listing \ref{list:oscillation_leapfrog} demonstrates how it is simple, clear and concise to solve a IVP by FOODiE solvers. Moreover, it proves how it is simple and effective to apply different solvers in a coupled algorithm, that greatly simplify the development of new hybrid solvers for self-adaptive time step size.

>>>>

\subsubsection{FOODIE aware implementation of an oscillation numerical solver}

The IVP \ref{eq:oscillation} can be easily solved by means of FOODIE library. The first block of a FOODIE aware solution consists to define an \emph{oscillation integrand field} defining a concrete extension of the FOODIE \emph{integrand} type. Listing \ref{list:oscillation_type} reports the implementation of such an integrand field that is contained into the tests suite shipped within the FOODIE library.

@@

As the listing \ref{list:oscillation_type} shows, the FOODIE implementation strictly corresponds to the mathematical formulation embracing all the relevant mathematical aspects into one derived type, a single \emph{object}. Here we not review the implementation of all deferred methods, this being out of the scope of the present work: the interested reader can see the tests suite sources shipped within the FOODIE library. However, some of these methods are relevant for our discussion, thus they are reviewed.

@@

Probably, the most important methods for an IVP solver is the residuals function. As a matter of facts, the ODE equations are implemented into the residuals function. However, the FOODIE ADT strongly alleviates the subtle problems that could arise when the ODE solver is hard-implemented within the specific ODE equations. As a matter of facts, the \emph{integrand} ADT specifies the precise interface the residuals function must have: if the client code implements a compliant interface, the FOODIE solvers will work as expected, reducing the possible errors location into the ODE equations, having designed the solvers on the ADT and not on the concrete type.

@@

The assignment overloading of the \emph{oscillation} type is the last key-method that enforces the conciseness of the FOODIE aware implementation. Listing \ref{list:oscillation_assign} reports the implementation of the assignment overloading. Essentially, to all the data members of the left-hand-side are assigned the values of the corresponding right-hand-side. Notably, for the assignment of the state vector and of the previous time steps solution array we take advantage of the automatic re-allocation of the left-hand-side variables when they are not allocated or allocated differently from the right-hand-side, that is a Fortran 2003 feature. In spite its simplicity, the assignment overloading is a key-method enabling the usage of FOODIE solver: effectively, the assignment between two \emph{integrand} ADT variables is ubiquitous into the solvers implementations, see equation \ref{eq:RK} for example.

@@

\paragraph{FOODIE numerical integration}

Using the above discussed \emph{oscillation} type it is very easy to solve IVP \ref{eq:oscillation} by means of FOODIE library. Listing \ref{list:oscillation_leapfrog} presents the numerical integration of system \ref{eq:oscillation} by means of the Leapfrog RAW-filtered method. In the example, the integration is performed with $10^4$ steps with a fixed $\Delta t=10^2$ until the time $t=10^6$ is reached. The example shows also that for starting a multi-step scheme such as the Leapfrog one a lower-oder or equivalent order one-scheme is necessary: in the example the first 2 steps are computed by means of one-step TVD/SSP Runge-Kutta 2-stages schemes. Note that the memory registers for storing the Runge-Kutta stages and the RAW filter displacement must be handled by the client code. Listing \ref{list:oscillation_leapfrog} demonstrates how it is simple, clear and concise to solve a IVP by FOODIE solvers. Moreover, it proves how it is simple and effective to apply different solvers in a coupled algorithm, that greatly simplify the development of new hybrid solvers for self-adaptive time step size.

### 2015-10-19 papers/announcement/src/tests/oscillation/oscillation.tex
Change project name to FOODIE
FOODIE => Fortran Object-Oriented Differential-equations Integration Environment

Why:

Due to a great idea of Damian.

This change addresses the need by:

Change all references to old FOODiE acronym.

Side effects:

A lot...

Because the Oscillation system \ref{eq:oscillation} posses a closed exact solution, the discussion on this test has twofolds aims: to assess the accuracy of the FOODiE's built-in solvers comparing the numerical solutions with the exact one and to demonstrate how it is simple to solve this prototypical problem by means of FOODiE.

>>>>

Because the Oscillation system \ref{eq:oscillation} posses a closed exact solution, the discussion on this test has twofolds aims: to assess the accuracy of the FOODIE's built-in solvers comparing the numerical solutions with the exact one and to demonstrate how it is simple to solve this prototypical problem by means of FOODIE.

### 2015-10-19 papers/announcement/src/tests/tests.tex
Change project name to FOODIE
FOODIE => Fortran Object-Oriented Differential-equations Integration Environment

Why:

Due to a great idea of Damian.

This change addresses the need by:

Change all references to old FOODiE acronym.

Side effects:

A lot...

For the assessment of FOODiE capabilities three different tests are discussed:

@@

With the first two pure ODEs tests we assess the accuracy of the FOODiE built-in solvers and demonstrate how to use the library in real applications. The last test discussion is mainly devoted to the demonstration of FOODiE application into a real, complex PDEs problem.

>>>>

For the assessment of FOODIE capabilities three different tests are discussed:

@@

With the first two pure ODEs tests we assess the accuracy of the FOODIE built-in solvers and demonstrate how to use the library in real applications. The last test discussion is mainly devoted to the demonstration of FOODIE application into a real, complex PDEs problem.
